Efficient and Accurate Modeling of Flow through Sedimentary Structures with MODFLOW 6

### 2024-06-10 latex/article.tex

\title{Efficient and Accurate Modeling of Flow through Dipping Aquifers with MODFLOW 6}

@@

\textbf{Article impact statement: Groundwater flow through dipping aquifers can be accurately simulated using a vertically offset grid combined with an unstructured grid approach and the XT3D multi-point flow formulation in MODFLOW 6.}

\begin{abstract}
Groundwater flow through dipping hydrogeologic layers is often modeled with MODFLOW by specifying top and bottom elevations of model layers to coincide with top and bottom elevations of hydrogeologic layers.  With this approach, adjacent cells within a model layer are vertically offset from one another.  While the standard formulation in MODFLOW does not rigorously account for these offsets in the underlying flow calculations, results are generally expected to be accurate provided dip angles do not exceed $10^{\circ}$  {\color{red} (KB: I would remove this sentence as even 10 degrees if anisotropic say 10:1 is still  a 50\% error)}.  To improve the accuracy of flow simulations on unstructured grids, and to allow accurate simulation of arbitrarily oriented two- or three-dimensional anisotropy, the optional XT3D multi-point flow formulation was introduced in MODFLOW 6.  The XT3D flow formulation is designed to account for geometric irregularities in the grid, including vertical offsets between laterally adjacent cells, and to provide accurate results for both isotropic and anisotropic groundwater flow.  \cite{bardot2022} showed, however, that neither the standard formulation nor the XT3D formulation produced the correct flow direction and magnitude in simple tests involving flow through a steeply dipping, permeable hydrogeologic layer.  In this paper, we explain that the inability of XT3D to give accurate results in the steeply dipping test problem of \cite{bardot2022} was caused by the limited cell connectivity that is inherent in the most commonly used discretization packages in MODFLOW 6. The DIS and DISV discretization packages are based on layered connectivity in which cells have lateral connections only with cells in the same grid layer.  We demonstrate that layered lateral connectivity is insufficient to allow accurate simulation of flow through a steeply dipping hydrogeologic layer on a vertically offset grid.  When lateral connections between cells in adjacent grid layers are included using the fully unstructured (DISU) grid type, XT3D is able to calculate the flow magnitude and direction accurately for both isotropic and anisotropic conductivity oriented along the permeable hydrogeologic layer.  To realize this improved accuracy, the permeable hydrogeologic layer must be discretized vertically using at least two model layers. {\color{red} (AMP: Given the idealized nature of the test problem, how general can we be with our conclusions? I pulled it back here relative to what you had originally, CDL, but left a slightly stronger statement in the Conclusions, at least for now.})  Our findings suggest that, given appropriate discretization and cell connectivity, the combination of vertically offset grids with the XT3D flow formulation in MODFLOW 6 can provide a computationally efficient and accurate method for simulating flow through isotropic and anisotropic steeply dipping layers, a capability that was previously reserved to finite-element-based simulators. {\color{red} (AMP: Any Pal and Edwards-type MPFA stuff that's relevant to that final phrase? See also Conclusions)}

{\color{red} Here's another option for a simpler abstract! }

	Dipping aquifers are often modeled using a deformed grid whereby each hydrogeological layer is represented using a single grid layer. However, until the release of MODFLOW 6, accurate modeling of groundwater flow through dipping anisotropic layers has not been possible due to limitations of its finite difference conductance-based solution. The introduction of the XT3D multi-point flow formulation has allowed accurate simulation of the full conductivity tensor, as well as handling the flow solution for irregular cell geometries for unstructured grids. Through the use of a benchmark, \cite{bardot2022} showcased MODFLOW 6's capability to model flow through complex sedimentary structures, previously restricted to finite element simulators. However, the benchmark study also brought to light the inadequacy of the deformed vertically offset grid for accurate modeling of dipping layers, with or without XT3D. This paper explains that the limitation of the deformed grid is a result of poor cell connectivity between vertically adjacent model layers. Nonetheless, we also demonstrate that dipping layers may be efficiently and accurately modeled using a deformed vertically offset grid with full-connectivity, which can be achieved by combining MODFLOW 6's DISU discretization package with the XT3D flow formulation. A repeat benchmark study shows that this method produces accurate head and flux results in isotropic and anisotropic dipping layers, providing a minimum of two grid layers per hydrogeological layer are deployed which allows for sufficient vertical connectivity.

@@

MODFLOW is commonly used to simulate groundwater flow in a wide variety of geological settings including sedimentary, alluvial and fractured rock aquifers \cite{modflow2005,Kumar2019}. Accurate simulation of the hydrostratigraphic sequences encountered in these geological settings requires that the model grid discretisation and solution technique can represent the scale and effects of important geologial features \cite{Reilly2004,bardot2022}, where permeability tensors are aligned along key structures \cite{hoaglund2003}. MODFLOW versions up to and including MODFLOW-2005 employ a structured grid based on a fixed number of layers, rows and columns \citep{modflow2005,modflow84,modflow88}. Further, through assuming that the hydraulic conductivity tensor aligns with the grid coordinates, the standard finite difference approximation can be applied accurately and efficiently to simulate groundwater flow \citep{modflow84}. The main trade-off in this approximation is that important hydrological and geological features are required to be mapped to the structured grids and orinetations need to be aligned to Cartesian coordinates. These requirements can result in pixelated features in plan view and, more significantly, create problems where geological layers are discontinuous at unconformities and faults, and where they are dipping and misaligned with the model grid \citep{bardot2022}.

In meeting the requirements for the standard finite difference simulation, two methods have traditionally been used to represent dipping geological features in MODFLOW in versions up to and including MODFLOW-2005. One approach is to map the geological features to a rectilinear grid \citep{modflow84} in a method referred to as ‘grid overlay’  \citep{hoaglund2003}. Grid overlay preserves the control volume finite difference approximation but may require a high level of vertical discretization to accurately represent variable layering \citep{Zyvoloski2006}. An alternative approach is to set each hydrostratigraphic unit as its own model layer, resulting in 'deformed layers' \cite{anderson2015applied,Reilly2004} in a 'vertically offset' or 'conforming finite difference grid' \citep{Zyvoloski2006}. When the layers in the offset grid are not strictly horizontal and of a constant thickness, the standard finite difference approximation is violated \citep{Zheng1994,anderson2015applied,Reilly2004}. To maintain a reasonable level of accuracy, \cite{anderson2015applied} recommend a $10^{\circ}$ dip as a practical upper limit for using vertical offsets.

The capability to represent complex ‘un-layered’ geology using unstructured grids was developed in versions of MODFLOW after MODFLOW 2005, namely MODFLOW-USG \citep{modflowusg} and MODFLOW 6 \citep{modflow6software}. Unstructured versions offered greater flexibility in representing the shape of geological features in plan view. To overcome the violations of the control volume finite difference approximations bought about by unsturctured grids, Ghost Node corrections were implemented in MODFLOW-USG \citep{modflowusg}, with MODFLOW 6 incorporating the XT3D option, which solves the full tensorial version of Darcy’s law \citep{modflow6xt3d}. Two of the discretisation options in MODFLOW 6 (known as the DIS and DISV options) preserve the layered connectivity approach from earlier MODFLOW versions \citep{modflow6framework}; however, both MODFLOW-USG, and the DISU option within MODLFOW 6 offer the capability to hydraulically connect multiple cells horizontally to the same face using the vertically staggered option with full connectivity, hence revolutionising the traditional layered-approach. We note that this type of connection is not implicit within the unstructured versions of the code, but needs to be defined by the user.

Recently Bardot et al. (2022) revisited the capacity of MODFLOW 6 to represent non-orthogonal geologic structures. A simple test problem was created using a highly conductive channel misaligned with a Cartesian grid in both plan and transect to test the code's capability to solve flow using flexible gridding in plan, and dipping anisotropic conductivity in transect. Through comparison with an analytical solution, the authors demonstrated that the application of XT3D produced excellent results in plan view with flexible gridding, highlighting the improved capabilities of MODFLOW 6. However, when attempting to represent an analytically identical problem in transect view, the authors experienced large errors in flow magnitudes and directions when using a vertically offset grid both with and without the use of XT3D. They recommended the use of the grid overlay method in conjunction with XT3D for simulating steeply dipping layers. A preliminary analysis by two authors of the present work (Provost and Langevin) hypothesized that the limited cell connectivity offered by the vertically offset grid was inadequate for allowing an accurate flow solution as it does not consider full connectivity between the offset layers.

Although the grid overlay method accurately simulates groundwater flow in dipping layers, the high vertical resolution required to represent geological structure becomes increasingly significant as dip increases, making the solution computationally prohibitive in some cases due to the large number of cells. This paper investigates the limitations of vertically offset grids and develops a solution for accurate and efficient grid representation in the vertical direction. We present a theoretical background which explains in detail the connectivity issues associated with vertically distorted grids, and propose a method to upgrade its accuracy using full connectivity grids together with XT3D option. The effectiveness of the proposed approach is validated through the benchmark problem of \cite{bardot2022}. Finally, we discuss the implications of our findings for practical groundwater problems involving steeply dipping hydrogeologic layers.

\section{Theoretical Background}

@@

Figure \ref{fig:schem-conn-area-flux}a shows a group of four cells in a two-grid-layer, cross-sectional MODFLOW 6 model of fully saturated flow through a permeable hydrogeologic layer (aquifer) with impermeable top and bottom boundaries. The cells have horizontal tops and bottoms and can therefore follow the dip of the hydrogeologic layer only on average. If the grid is represented using a layered grid type (DIS or DISV) in MODFLOW 6, a cell is hydraulically connected to each adjacent cell in the same grid layer, regardless of whether or not the cells overlap, and with each overlying and underlying cell, as indicated by the blue lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a. In this work we call this a vertically offset grid with ``layered connectivity.''  Note that cells that have overlapping vertical faces but are in different grid layers (e.g., cells 1A and 2B) do not have a direct hydraulic connection in a grid with layered connectivity. The addition of such connections between cells in different grid layers, as indicated by the red lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a, results in a grid with ``full connectivity.''  In MODFLOW 6, only layered connectivity can be represented in structured (DIS) and vertex-based (DISV) grids. Fully unstructured (DISU) grids allow representation of any kind of cell connectivity, including layered or full connectivity.

The default method for representing the flow between two hydraulically connected model cells in MODFLOW 6 is the ``conductance-based flow formulation'' \citep{modflow6gwf}. In this formulation, which is based on the commonly used two-point flux approximation which arises from the starndard finite difference formulation \citep{modflow84}, the flow is proportional to the difference in the hydraulic heads computed at the two cell centers and a hydraulic conductance based on an effective hydraulic conductivity for the connection between the cells. On strictly rectilinear grids, such as structured MODFLOW grids with no vertical offset between cells in the same grid layer, and hydraulic conductivity that is isotropic or has anisotropy aligned with the three mutually perpendicular grid directions, the conductance-based formulation is second-order accurate \citep{dehotin2010modeling, modflow6gwf}. This is because strictly rectilinear grids satisfy the ``control-volume finite-difference (CVFD) requirement'' that the straight-line connection between two cell centers must intersect the midpoint of the cell-cell interface at a right angle \citep{narasimhan1976integrated}. However, vertically offset grids violate the CFVD requirement because nominally ``horizontal'' connections between cell centers in the same grid layer are not strictly horizontal and therefore not perpendicular to the vertical cell-cell interfaces they intersect. For example, in Figure \ref{fig:schem-conn-area-flux}a, the connection between cells 1A and 1B is not perpendicular to the cell-cell interface, which compromises the accuracy of the conductance-based formulation.

The conductance-based formulation was the only flow formulation available in versions of MODFLOW prior to MODFLOW 6. As mentioned earlier, the limitations of the conductance-based formulation on vertically offset grids were well recognized. However, the ability to use unstructured grids in MODFLOW-USG and MODFLOW 6 introduced new ways to violate the CVFD requirement, and thereby render the conductance-based formulation less accurate, even in the absence of vertical offsets. The concept of a ``ghost node'' was included in MODFLOW-USG \citep{modflowusg}, and subsequently in MODFLOW 6, as a simple and optional way to improve the accuracy for grids that violated the CVFD requirement.  However, implementation of the ghost-node correction is highly problem dependent, and it is not obvious how to specify optimal ghost-node locations and weights on complex, unstructured grids.  As a more ``automatic'' alternative to the ghost-node method for improving accuracy, and as a way to allow accurate simulation of arbitrarily oriented two- or three-dimensional anisotropy, the optional XT3D flow formulation \citep{modflow6xt3d} was introduced in MODFLOW 6.  XT3D formulates the flow between two cells by interpolating head values from the two cells and their neighboring cells to construct an estimate of the full, two- or three-dimensional head-gradient vector on each side of the cell-cell interface; applying the cell conductivity tensors to obtain an estimate of the groundwater flux  normal to the interface on each side of the interface; and reconciling the two flux estimates to ensure continuity of flow across the interface. By performing flow calculations using Darcy's Law in its tensorial form, based on a ``multi-point'' approximation of the head gradient, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid.

The dipping-hydrogeologic-layer benchmark problem of \cite{bardot2022} used a vertically offset layered grid, which is the type of grid used most often in MODFLOW models. Test simulations using the conductance-based flow formulation for a steeply dipping hydrogeologic layer (which we will call the ``aquifer'') embedded in a near-impermable domain yielded significant errors in the simulated flows. The simulated groundwater flux in the middle of the aquifer was not along the $30^{\circ}$ incline of the aquifer as expected, but nearly horizontal ($0.03^{\circ}$ incline), and its magnitude was overestimated by 15\%. Significant errors in the simulated flows were expected in this case, given that the conductance-based flow formulation does not rigorously account for violations of the CVFD requirement that the cell-cell connection be perpendicular to the cell-cell interface, and that it takes the connection length to be the horizontal distance between cell centers.  Given the ability of XT3D to account rigorously for cell connections that are not perpendicular to cell-cell interfaces, and for the increase in connection length due to the slope of the connection, rerunning the simulation with XT3D activated could have been expected to improve the flow solution substantially. However, \cite{bardot2022} observed similarly poor results with XT3D: the flux was still nearly horizontal ($0.01^{\circ}$ incline), and its magnitude was overestimated by 12\%. Based on a preliminary analysis by two authors of the present work (Provost and Langevin), \cite{bardot2022} hypothesized that the unexpectedly large error in the flow solution with XT3D was related to inadequate cell connectivity.

The role of cell connectivity in enabling accurate simulation of flow along a dipping hydrogeologic layer can be understood by considering Figure \ref{fig:schem-conn-area-flux}b. Components of the groundwater flux (specific discharge) vector are displayed on each cell-cell interface and are representative of steady, uniform flow along the hydrogeologic layer. Gray vectors represent the uniform flux oriented along the hydrogeologic layer, which the model is attempting to simulate. Blue vectors represent the flux components normal to cell-cell interfaces that correspond to blue connections in Figure \ref{fig:schem-conn-area-flux}a, which are common to both the layered-connectivity and full-connectivity grids. Red vectors represent the flux components normal to cell-cell interfaces that correspond to red connections in Figure \ref{fig:schem-conn-area-flux}a, which are lacking in the layered-connectivity grid.

@@

@@

To simulate flow along an aquifer of uniform thickness inclined at angle $\theta$ relative to the horizontal ($30^{\circ}$ for the base case), heads are specified at the centers of cells along the perimeter of the model using a formula that corresponds to a uniform, unit head gradient:
\begin{equation}
\label{eqn:head_analyt_along}
h = - x \cos \theta - z \sin \theta.
\end{equation}

@@

Two methods of discretization are used to simulate the test problem: (1) vertically offset grid of the MODFLOW 6 DIS (structured) type, and (2) vertically offset grid of MODFLOW 6 DISU (unstructured) type. The cell geometry in these two discretization methods is identical, but the cell connectivity is different. The DIS grid uses layered connectivity, and the DISU grid includes the additional connections between cells in adjacent grid layers needed to realize full connectivity. The DISU grid is constructed by starting with a DIS grid, converting it to an equivalent DISU grid, and then implementing full connectivity. The base case uses grids consisting of 11 columns and 9 layers of cells, with 3 layers within the aquifer, 3 layers in the domain above the aquifer, and 3 layers in the domain below the aquifer.  The model width was set to 11 m such that each column was 1m wide. Cells within the aquifer were chosen to stay square (i.e. 1 x 1 m) for the purpose of the test problem to eliminate distortion effects. Corner cells in the lower left and upper right corner were also set to be the same dimension as the aquifer cells. The resulting base model vertical height is 14.77 m.

@@

Jupyter notebooks created for the test problem are available in the Supporting Information that accompanies this paper. The notebooks leverage the capabilities of FloPy \citep{bakker2016scripting, hughes2023flopy} to assist in setting up input for and processing results from the MODFLOW 6 simulations. A Python script for converting structured, DIS grids to unstructured, DISU grids with full connectivity is included.

@@

To demonstrate the effect of cell connectivity on the accuracy of simulated flow through a steeply dipping hydrogeologic layer (aquifer), results for the base case described in the previous section are compared for simulations that use either layered or full connectivity, and either the standard or the XT3D flow formulation. The base case is then modified to further investigate the effects of grid resolution, aquifer dip angle, conductivity contrast between the aquifer and the surrounding domain, and anisotropic hydraulic conductivity.

Plots of specific discharge (groundwater flux) at cell centers and flows across cell-cell interfaces (face flows) calculated by MODFLOW 6 are used to illustrate the accuracy of the simulated flows. The magnitude and direction of specific discharge at the cell in the center of the aquifer, at which boundary effects are presumably minimized, and the total volumetric flow rate entering the aquifer through the left-hand boundary, which consists of constant-head (CHD) cells, are also reported as ``representative'' values of flow. Maximum head errors are also reported for the base case.

@@

The numerical solution for the layered connectivity grid confirms the findings of \cite{bardot2022} in that fluxes are predominantly horizontal (top row, Figure \ref{fig:fig2}). We see that despite an enforced hydraulic gradient of $30^{\circ}$, the vertical component of the flux cannot propagate through the aquifer given that cells along the aquifer top are hydraulically disconnected from the low permeability domain, hence forcing flow horizontally (red arrows on right panel). The numerical solution for this grid type is a problem of cell connectivity, and thus the XT3D formulation does not resolve this issue (second row, Figure \ref{fig:fig2}). Results for the full-connectivity grid with the standard conductance-based formulation shows an improvement in the flow solution (third row, Figure \ref{fig:fig2}), with the XT3D formulation successfully reproducing the analytical solution (bottom row, Figure \ref{fig:fig2}). We see that the grid with full connectivity allows incorporation of the vertical flow component, thus permitting cross-connection between grid layers.

The results shown in Figure \ref{fig:fig2}a and b , which are based on a vertically offset grid with layered connectivity, are consistent with the findings of \cite{bardot2022}. Despite boundary heads designed to induce specific discharge of 1 m/d along the $30^{\circ}$ incline of the aquifer, the simulated flow within the aquifer is predominantly horizontal, and the magnitude of the specific discharge at the center of the aquifer and the total flow entering the aquifer are both overestimated. The XT3D flow formulation is unable to compensate for the inadequate cell connectivity (Figure \ref{fig:fig2}b), which does not allow flow between laterally adjacent cells in different grid layers.

The vertically offset grid with full connectivity (Figure \ref{fig:fig2}c and d) show significantly improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c). Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d).

\begin{figure}[p!]
	\begin{center}
	\includegraphics[scale=0.8]{../figures/fig2_paper.png}
	\caption{Numerical results for the test problem using base-case settings for layered-connectivity and full-connectivity grids, with the standard conductance based formulation as well as the XT3D formulation. The left panel of each scenario shows the calculated specific discharge at each cell centre within the aquifer (black arrows) and domain (grey arrows). The middle panel shows the face flows (red arrows). The right panel shows head contours for the analytical solution (black) as well as the numerical solution (blue).}
	\label{fig:fig2}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig3paper.png}
	\caption{Graphs indicating how increasing the number of grid layers within a single hydrogeologic layer (in this case, the aquifer) affects flux magnitude of middle cell (top), flux direction of middle cell (middle) and volumetric flow through the hydrogeologic layer (bottom). We immediately notice the errors layered-connectivity grids incur (left panel), and how using a full-connectivity grid (right panel), particularly in conjunction with XT3D (orange line) reproduces the analytical solution from two grid layers used to discretize the hydrogeologic layer.}
	\label{fig:fig3}
	\end{center}
\end{figure}

>>>>

%\title{Improving Accuracy of Simulated Flows in Steeply Dipping Layers Using Full-Connectivity Grids with the XT3D Multi-Point Flux Approximation in MODFLOW 6}

% suggested title revision
\title{Efficient and Accurate Modeling of Flow through Sedimentary Structures with MODFLOW 6}

%\title{Efficient and Accurate Modeling of Flow through Dipping Aquifers with MODFLOW 6}

\title{The Importance of Cell Connectivity for Efficient and Accurate Modeling of Flow through Dipping Aquifers with MODFLOW 6}

@@

%\textbf{Article impact statement: Groundwater flow through dipping aquifers can be accurately simulated using a vertically offset grid combined with an unstructured grid approach and the XT3D multi-point flow formulation in MODFLOW 6.}

\textbf{Article impact statement:} Accurate simulation of groundwater flow through dipping aquifers using MODFLOW 6 can require enhanced cell connectivity and use of a multi-point formulation for flow between cells. This can be achieved using an unstructured grid and the XT3D flow formulation.

\begin{abstract}

Groundwater flow through a dipping aquifer is often modeled using a single layer of model cells to represent the aquifer. The layer of cells is ``deformed'' to follow the top and bottom elevations of the aquifer. When this approach is used in MODFLOW, adjacent cells within the model layer are vertically offset from one another, and the standard conductance-based (two-point) formulation for flow between cells does not rigorously account for these offsets. The XT3D multi-point flow formulation introduced in MODFLOW 6 is designed to account for geometric irregularities in the grid, including vertical offsets, and to provide accurate results for both isotropic and anisotropic groundwater flow.  A recent study evaluated the performance of the standard formulation and XT3D using a simple, synthetic benchmark model of a steeply dipping aquifer. Although XT3D generally improved the accuracy of flow simulations relative to the standard formulation as expected, neither formulation produced accurate flows in cases that involved large vertical offsets. In this paper, we explain that the inability of XT3D to produce accurate flows in the steeply dipping aquifer benchmark was not due to an inherent limitation of the flow formulation, but to the limited cell connectivity inherent in the most commonly used discretization packages in MODFLOW 6. Furthermore, we demonstrate that XT3D is able to produce the expected accuracy when adequate cell connectivity is introduced using MODFLOW's unstructured grid type and the aquifer is discretized vertically using at least two model layers.

@@

Numerical simulation of layered aquifer systems can involve a tradeoff between efficiency and accuracy. A traditional approach to discretizing layered aquifer systems, called the ``grid overlay'' method by \cite{hoaglund2003}, maps the heterogeneous hydrologic properties onto a rectilinear grid \citep{modflow84}. While this method preserves the benefits of the rectilinear grid structure, it can require a large number of model grid layers to adequately discretize hydrogeogic layers of variable dip and thickness \citep{Zyvoloski2006}.

An alternative approach, called the ``boundary-matching" method by \cite{hoaglund2003}, allows model grid layers to "deform" vertically to follow the undulations of hydrogeologic layers. This method ``minimize[s] the number of model [grid] layers required to simulate an aquifer system'' \citep{modflow84}. In fact, it is common practice to represent each hydrogeologic layer using a single, boundary-matching model grid layer. However, the non-orthogonality of the resulting grid can be a source of significant model error \citep{hoaglund2003}.

In models based on the Control Volume Finite Difference method {(\color{red} ref for this?)}, which is the method used in the MODFLOW family of hydrologic simulation codes \citep{modflow6gwf}{\color{red} (ref to our new MODFLOW overview paper!)}, the loss of accuracy associated with the boundary-matching approach results from the use of a "two-point" formulation to calculate flows between adjacent grid cells. In the two-point formulation, the flow between two cells is proportional to the difference between the heads calculated in those two cells. For simulations of flow through isotropic porous media, the two-point flow formulation is most accurate when the grid satisfies the so-called ``CVFD requirement" \citep{modflow6gwf} that the straight-line connection between two cell centers must intersect the midpoint of the cell interface at a right angle \citep{narasimhan1976integrated}. However, boundary-matching grids violate the CVFD requirement at vertical interfaces between cells in the same model layer because the corresponding cell connections are not strictly horizontal.

In MODFLOW, the top and bottom surfaces of model cells are horizontal \citep{modflow6gwf}. Thus, when boundary matching is implemented in a two-dimensional (2D) cross-sectional or three-dimensional (3D) MODFLOW model, the tops of laterally adjacent cells are vertically offset from each other (as are the cell bottoms) and can follow the undulations of the aquifer boundaries only on average. Following \cite{bardot2022}, we call this type of grid ``vertically offset." In a vertically offset grid, subhorizontal cell connections across vertical cell interfaces violate the CVFD requirement, which reduces the accuracy of two-point flow formulation. For an isotropic aquifer, \cite{anderson2015applied} recommend a $10^{\circ}$ dip as a practical upper limit when using the two-point flow formulation. {\color{red} (AMP: Was this offered in the context of MODFLOW, or more generally??? I don't have that book.)}

The introduction of unstructured grids into MODFLOW \citep{modflowusg, modflow6gwf} presented new ways for a grid to violate the CVFD requirements. To help overcome the limitations of the two-point flow formulation on geometrically irregular grids, MODFLOW-USG and MODFLOW 6 offer a Ghost-Node Correction (GNC) Package \citep{modflowusg, modflow6gwf}{\color{red} (ref to Dickenson and others 2007 ?)} However, configuration of optimal ghost node locations and interpolation weights is highly problem-dependent, and we are not aware of guidelines for handling general unstructured grids or arbitrarily oriented 3D anisotropy.

The optional XT3D flow formulation \citep{modflow6xt3d} introduced in MODFLOW 6 offers a more ``automatic" alternative to ghost nodes. XT3D is a multi-point flow formulation {\color{red} (refs for multi-point)} based on gradient reconstruction {\color{red} (refs for gradient reconstruction)}. By performing flow calculations using Darcy's Law in its tensorial form using a multi-point approximation of the head gradient vector, XT3D accounts for both arbitrarily oriented anisotropy and geometric irregularity of the grid. In theory, XT3D should be able to compensate for the geometric irregularity introduced by vertical offsets between adjacent cells in the same grid layer.

\cite{bardot2022} tested the ability of MODFLOW 6 to accurately simulate flow through sedimentary structures, both with and without XT3D. The simplest 2D cross-sectional (``transect") version of their benchmark model simulated flow through a steeply dipping, homogeneous, isotropic aquifer of uniform thickness embedded between two confining units. In test runs that used the boundary-matching method on a vertically offset grid, both the two-point flow formulation and XT3D produced flows that differed markedly from the expected magnitude and direction throughout the aquifer, regardless of how finely the aquifer was discretized. They noted, however, that a preliminary analysis by two authors of the present work (Provost and Langevin) suggested that the limited cell connectivity offered by the layered, vertically offset grids was inadequate to allow an accurate flow solution.

To our knowledge, the implications of limited cell connectivity in layered, vertically offset grids for the accuracy of simulated flows, and the potential to improve accuracy using enhanced connectivity between model layers, have not been previously recognized or investigated. In this paper, we present theoretical explanations for why ``layered" connectivity in a vertically offset grid is inadequate for accurate simulation of flows in steeply dipping hydrogeologic layers, and why using what we call ``full" cell connectivity together with XT3D should improve accuracy. Then, in a set of test problems similar to those of \cite{bardot2022}, we demonstrate the effectiveness of using full connectivity with XT3D. Finally, we discuss the implications of our findings for practical groundwater problems involving steeply dipping hydrogeologic layers.

\section{Theory}

Versions of MODFLOW prior to MODFLOW-USG \citep{modflowusg} supported only structured grids composed of rows, columns, and layers of cells, which could be vertically offset. This grid type, called ``DIS," continues to be supported in MODFLOW-USG and MODFLOW 6, which also offer an unstructured grid type called ``DISU" \citep{modflowusg, modflow6gwf}. In addition, MODFLOW 6 offers an  unstructured grid type called ``DISV," which is based on specification of cell vertices \citep{modflow6gwf}. The DIS and DISV grid types are based on model layers, with the same horizontal grid structure within each layer. The DISU grid type is more flexible and does not depend on the concept of model layers.

@@

Figure \ref{fig:schem-conn-area-flux}a illustrates what we refer to as "layered" and "full" cell connectivity. The focus is on a group of four cells in a two-grid-layer, vertically offset MODFLOW 6 model of fully saturated flow through a permeable hydrogeologic layer (aquifer) with impermeable top and bottom boundaries. If the grid is represented using a layered grid type (DIS or DISV), the grid has what we call ``layered connectivity": a cell is hydraulically connected to each laterally adjacent cell in the same grid layer (regardless of whether or not the cell faces actually overlap), and with each overlying and underlying cell, as indicated by the blue lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a. In layered connectivity, two cells that have overlapping vertical faces but are in different grid layers (e.g., cells 1A and 2B) notably do not have a direct hydraulic connection. The addition of such connections between cells in different grid layers, as indicated by the red lines that connect cell centers in Figure \ref{fig:schem-conn-area-flux}a, results in a grid with what we call ``full connectivity.''  Implementation of full connectivity in MODFLOW 6 requires that the vertically offset grid be represented using the fully unstructured (DISU) grid type.

The role of full cell connectivity in enabling accurate simulation of flow along a dipping hydrogeologic layer can be understood by considering Figure \ref{fig:schem-conn-area-flux}b. Components of the groundwater flux (specific discharge) vector are displayed on each cell-cell interface and are representative of steady, uniform flow along the hydrogeologic layer. Gray vectors represent the uniform flux oriented along the hydrogeologic layer, which the model is attempting to simulate. Blue vectors represent the flux components normal to cell-cell interfaces that correspond to blue connections in Figure \ref{fig:schem-conn-area-flux}a, which are common to both the layered-connectivity and full-connectivity grids. Red vectors represent the flux components normal to cell-cell interfaces that correspond to red connections in Figure \ref{fig:schem-conn-area-flux}a, which are lacking in the layered-connectivity grid.

@@

Note how the vertically offset grid violates the CFVD requirement because nominally ``horizontal'' connections between cell centers in the same grid layer are not strictly horizontal and therefore not perpendicular to the vertical cell-cell interfaces they intersect. For example, in Figure \ref{fig:schem-conn-area-flux}a, the connection between cells 1A and 1B is not perpendicular to the cell-cell interface, which reduces the accuracy of the conductance-based formulation.

@@

To simulate flow along an aquifer of uniform thickness inclined at angle $\theta$ relative to the horizontal ($30^{\circ}$ for the base case), heads are specified at the centers of cells along the perimeter of the model using a formula that corresponds to a uniform, unit head gradient:
\begin{equation}
\label{eqn:head_analyt_along}
h = - x \cos \theta - z \sin \theta,
\end{equation}

@@

The test simulations use grids of MODFLOW 6 DISU (unstructured) type, which offers the flexibility to implement either layered or full connectivity. In all cases, a grid of MODFLOW 6 DIS (structured) type is first created and then converted a DISU grid. In simulations that use layered connectivity, the DISU grid is formulated to have the same cell geometry and connectivity as the original DIS grid. In simulations that use full connectivity, additional connections between adjacent layers are added during the conversion process. The base case uses grids consisting of 11 columns and 9 layers of cells, with 3 layers within the aquifer, 3 layers in the domain above the aquifer, and 3 layers in the domain below the aquifer.  The model is 11 m wide, and each column is 1 m wide. Cells within the aquifer are square (1 m x 1 m). Corner cells in the lower left and upper right corner were also set to be the same dimension as the aquifer cells. The resulting base model vertical height is 14.77 m.

@@

Jupyter notebooks created for the test problem are available in the Supporting Information that accompanies this paper {\color{red} (best place???)}. The notebooks leverage the capabilities of FloPy \citep{bakker2016scripting, hughes2023flopy} to assist in setting up input for and processing results from the MODFLOW 6 simulations. A Python script for converting structured, DIS grids to unstructured, DISU grids with full connectivity is included.

@@

To demonstrate the effect of cell connectivity on the accuracy of simulated flow through a steeply dipping aquifer, results for the base case described in the previous section are compared for simulations that use either layered or full connectivity, and either the two-point flow formulation (also known as the "standard conductance-based formulation" in MODFLOW 6) or the XT3D flow formulation. The base case is then modified to further investigate the effects of grid resolution, aquifer dip angle, conductivity contrast between the aquifer and the surrounding domain, and anisotropic hydraulic conductivity.

Plots of specific discharge (groundwater flux) at cell centers and flows across cell-cell interfaces (face flows) calculated by MODFLOW 6 are used to illustrate the accuracy of the simulated flows. The magnitude and direction of specific discharge at the cell in the center of the aquifer, at which boundary effects are presumably minimized, and the total volumetric flow rate entering the aquifer through the left-hand boundary, which consists of constant-head (CHD) cells, are also reported as representative values of flow. Maximum head errors are also reported for the base case.

@@

The numerical solution for the layered-connectivity grid confirms the findings of \cite{bardot2022} in that fluxes are predominantly horizontal (Figure \ref{fig:fig2}a). We see that despite an enforced hydraulic gradient of $30^{\circ}$, the vertical component of the flux cannot propagate through the aquifer given that cells along the aquifer top are hydraulically disconnected from the low permeability domain, hence forcing flow horizontally (red arrows on right panel). The numerical solution for this grid type is a problem of cell connectivity, and thus the XT3D formulation does not resolve this issue (Figure \ref{fig:fig2}b). Results for the full-connectivity grid with the standard conductance-based formulation shows an improvement in the flow solution (Figure \ref{fig:fig2}c), with the XT3D formulation successfully reproducing the analytical solution (Figure \ref{fig:fig2}d). We see that the grid with full connectivity allows incorporation of the vertical flow component, thus permitting cross-connection between grid layers.

The results shown in Figures \ref{fig:fig2}a and b , which are based on a vertically offset grid with layered connectivity, are consistent with the findings of \cite{bardot2022}. Despite boundary heads designed to induce specific discharge of 1 m/d along the $30^{\circ}$ incline of the aquifer, the simulated flow within the aquifer is predominantly horizontal, and the magnitude of the specific discharge at the center of the aquifer and the total flow entering the aquifer are both overestimated. The XT3D flow formulation is unable to compensate for the inadequate cell connectivity (Figure \ref{fig:fig2}b), which does not allow flow between laterally adjacent cells in different grid layers.

The vertically offset grid with full connectivity (Figures \ref{fig:fig2}c and d) shows significantly improved results, even for the standard flow formulation (Figure \ref{fig:fig2}c). Using the XT3D flow formulation provides the most accurate results, as expected (Figure \ref{fig:fig2}d).

\begin{figure}[p!]
	\begin{center}
	\includegraphics[scale=0.8]{../figures/fig2_paper.png}
	\caption{Numerical results for the test problem using base-case settings for layered-connectivity and full-connectivity grids, with the standard conductance-based formulation as well as the XT3D formulation. The left panel of each scenario shows the calculated specific discharge at each cell center (arrows). The middle panel shows the face flows (arrows). The right panel shows head contours for the analytical solution (black) as well as the numerical solution (blue).}
	\label{fig:fig2}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center}
	\includegraphics[scale=0.9]{../figures/fig3paper.png}
	\caption{Graphs indicating how the number of grid layers within the aquifer affects the flux magnitude in the center cell (top), the flux direction in center cell (middle) and the volumetric flow through the aquifer (bottom). Errors are minimized by using full connectivity (right column), the XT3D flow formulation (orange lines) and at least two grids layers in the aquifer.}
	\label{fig:fig3}
	\end{center}
\end{figure}
